Error Reporting Parsers: a Monad Transformer Approach

### 2013-05-27 article.tex
Comments
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

\subsection{Parser-specific combinators}
\verb+item+ is a very simple parser; it 
consumes a single token if available, and fails otherwise:
\begin{verbatim}
item :: Parser t t
item = Parser f
  where 
    f [] = Nothing
    f (x:xs) = Just (xs, x)
\end{verbatim}

Analagous to \verb+Control.Monad.mfilter+ is the \verb+check+ combinator, which is used to
validate a parse result:
\begin{verbatim}
check :: (Monad f, Alternative f) => (a -> Bool) -> f a -> f a
check p m =
    m >>= \x -> if p x then return x else empty
\end{verbatim}

@@

Parsers built out of \verb+literal+ only succeed when the next token matches 
the given one:
\begin{verbatim}
*Main> getParser (literal 'a') "abcde"
Just ("bcde",'a')

@@

*Main> getParser parseNest "(()abc(def)()"
Just ("(()abc(def)()",[])
\end{verbatim}
In the rest of this article, we'll rework and augment these basic parser 
combinators by genericizing the implementations and extending the result types,
while taking advantage of monad transformers to keep them modular.

\section{Monad transformers}
Monad transformers \cite{liang} allow monads to be modularly combined to form 
new, combined monads which support operations of each of their constituents.  
There are several useful monads for parsing, and by applying monad transformers, 
we can build semantically rich parsers without sacrificing modularity.
We'll use the standard transformers/mtl \cite{mtl} libraries -- 
they're easy to get, easy to use, and very useful.

First, note that the parser type we used earlier -- \verb+[t] -> Maybe ([t], a)+
-- can be built out of the StateT monad transformer applied to Maybe.  We can 
now define the new parser type as:
\begin{verbatim}
import Control.Monad.State       (StateT(..))

@@

The \verb+item+ parser is reimplemented using the combinators from MonadState --
token lists are the 'state' that \verb+item+ must inspect and modify:
\begin{verbatim}
{-# LANGUAGE   FlexibleContexts  #-}

@@

item :: (MonadState [t] m, Alternative m) => m t
item =
    get >>= \xs -> case xs of
                        (t:ts) -> put ts *> pure t;
                        []     -> empty;
\end{verbatim}
Note that the FlexibleContexts language extension is required by the 
MonadState constraint in the type signature.  This formulation of monad
transformers relies heavily on extensions to the type system, more of which 
we'll have to use for some of the later code in this article.

\section{Introducing Woof:  a Simple Lisp}
The simple language we'll use as a motivation for building parsers is Woof, a 
simple dialect of Lisp.  We'll be progressively adding features to a simple 
initial implementation on our way to creating an error-reporting parser.

@@

\subsection{Preliminaries}
Our first Woof parser will be responsible for: 1) determining whether input
text conforms to the language definition, and 2) building an 
Abstract Syntax Tree representing the structure of the parsed input.
Here's the AST definition we'll use:
\begin{verbatim}
data AST
    = ASymbol String
    | ALambda [String] [AST]
    | ADefine String AST
    | AApp    AST  [AST]
  deriving (Show, Eq)
\end{verbatim}

@@

And some examples to make sure it's behaving correctly:
\begin{verbatim}
*Main> map (runParser  opencurly) ["{abc", "}abc", "(abc", "abc"]
[Just ('{', "abc"), Nothing, Nothing, Nothing]
\end{verbatim}

@@

Of course not -- we haven't told it how to skip comments and whitespace yet!
Let's do that now, first by defining the whitespace and comment patterns, then
by creating a \verb+munch+ combinator that takes in a parser, 
throws away all leading junk, and finally runs the parser.  Also, we rename
the \verb+some+ and \verb+many+ functions from Control.Alternative to more
clearly indicate their semantics:
\begin{verbatim}
many0 = many
many1 = some

@@

On to Lambda.  Not only does Lambda have additional syntax with its parameter 
list, but we also need to ensure that the parameter names are unique.  We can 
do that using the \verb+check+ combinator again:
\begin{verbatim}
lambda = 
    opencurly                       *>
    check (== "lambda") symbol      *>
    opencurly                       *>
    pure ALambda                   <*>
    check distinct (many0 symbol)  <*>
    (closecurly                     *>
     many1 form                    <*
     closecurly)
  where
    distinct names = length names == length (nub names)
\end{verbatim}
If the symbols are distinct, the parameter list subparser will succeed,
whereas if there's a repeated symbol, it will fail.

@@

We also need to munch any trailing comments and whitespace -- otherwise \verb+end+
would fail -- so the final woof parser is:
\begin{verbatim}
woof = many1 form <* munch end
\end{verbatim}

@@

While our previous parsers had two different results -- Nothing and Just ... --
our new parsers have three possible results:  success, failure, and error.  Note
that successful and failing parses are wrapped in an additional \verb+Right+
constructor:
\begin{verbatim}
*Main> runParser item "abc"
Right (Just ('a',"bc"))

@@

\subsection{Reporting position}
To report error positionr, we'll pre-process the input
string to calculate the line and column number of each character.  Then instead 
of our parsers operating on a list of characters, they'll operate on a list of 
Tokens, where a Token is a character tupled up with a line number and column 
number, which we'll express with a simple typedef and some accessor functions:
\begin{verbatim}
type Token = (Char, Int, Int)
chr  (a, _, _)  =  a
line (_, b, _)  =  b
col  (_, _, c)  =  c
\end{verbatim}
The function responsible for calculating the position will simply transform a 
list of characters into a list of tokens:
\begin{verbatim}
countLineCol :: [Char] -> [Token]
countLineCol = reverse . snd . foldl f ((1, 1), [])
  where
    f ((line, col), ts) '\n' = ((line + 1, 1), ('\n', line, col):ts)
    f ((line, col), ts)  c   = ((line, col + 1), (c, line, col):ts)
\end{verbatim}

\subsection{Token parsers}
By adding the position to the input, we've created a new problem 
for ourselves: our original token parsers operated on 
character lists, but we now need them to work on token lists.  
The cause of the problem is the \verb+literal+ combinator, so we
replace it with a similar one for token lists:
\begin{verbatim}
character :: (MonadState [Token] m, 
              Alternative m) => Char -> m Token
character c = satisfy ((==) c . chr)
\end{verbatim}

And now our token parsers are:
\begin{verbatim}
whitespace = many1 $ satisfy (flip elem " \n\t\r\f" . chr)
comment = pure (:) <*> character ';' <*> many0 chars
  where chars = not1 $ character '\n'

@@

\subsection{Error messages}
Now we modify the remaining parsers to report accurate and useful errors.
Here are the error messages that we'll use to report each error:
\begin{verbatim}
eAppOper    =  "application: missing operator"
eAppClose   =  "application: missing close parenthesis"
eDefSym     =  "define: missing symbol"
eDefForm    =  "define: missing form"
eDefClose   =  "define: missing close curly"
eLamParam   =  "lambda: missing parameter list"
eLamDupe    =  "lambda: duplicate parameter names"
eLamPClose  =  "lambda: missing parameter list close curly"
eLamBody    =  "lambda: missing body form"
eLamClose   =  "lambda: missing close curly"
eSpecial    =  "special form: unable to parse"
eWoof       =  "woof: unparsed input"
\end{verbatim}

The MonadError type class provides two useful combinators, \verb+throwError+ for 
generating errors, and \verb+catchError+ for dealing with them.  We can ignore 
the second because we don't need to recover from errors, just report them, 
but when should an error be generated and what 
data should it include?  \verb+throwError+ is too basic for our needs, so we'll 
introduce the \verb+commit+ combinator:
\begin{verbatim}
commit :: (MonadError e m, Alternative m) => e -> m a -> m a
commit err p = p <|> throwError err
\end{verbatim}

This combinator takes two arguments: 1. an error value, and 2. a monadic 
computation.  It first tries to run the computation, and if the computation 
fails, generates an error whose contents are the first argument.  The 
definition of 'failure' is provided by the Alternative instance of \verb+m+.

What data do we want to include in our errors?  A String describing the error 
and a token reporting the position will describe the error decently:
\begin{verbatim}
type Error = (String, Token)
\end{verbatim}

@@

Note that \verb+commit+ isn't used until we can be sure that we're in the right 
rule:  if, in the \verb+define+ parser, we committed to parsing the "define" 
keyword, we wouldn't be able to backtrack and try the \verb+lambda+ parser if 
that failed.  Thus, if errors are treated as unrecoverable, it's important not to 
place a \verb+commit+ where backtracking might be necessary to parse a valid 
alternative.  However, in our example, once 
we see an open curly brace and a define, we're sure that we're in the \verb+define+
rule, and allowing useless backtracking would only destroy our ability to report 
errors cleanly and accurately.

@@

Incorrect input is reported as an error.  For each of the items in our error spec, 
we can show that our parser correctly reports an error: 
\begin{verbatim}
*Main> p "()"
Left ("application: missing operator",('(',1,1))

@@

\section{Stack Order}
Not only are the contents of a transformer stack important, but the order of the
layers as well \cite{stack}.  Let's revisit our first woof parser -- what if our 
stack had been Maybe/State instead of State/Maybe?
\begin{verbatim}
State/Maybe:   s -> Maybe (a, s)

@@

On the other hand, the order of Error and Maybe relative to each other isn't 
important.  But don't just take my word for it -- try it for yourself!

Another important point to note is that our parsers don't specify
which order -- Maybe/State or State/Maybe -- they prefer, and will 
work with either.  This means that the semantics are ultimately determined by 
the actual stack we use.

\section{Conclusion}
Monad transformers provide an elegant solution to error reporting in parser
combinators.  Better yet, they are extensible -- we could go back and add more
features to our parsers using additional transformers, perhaps extending our
parser to save partial results, so that if there's an error, it reports the
progress that it had made before the error.  We could do that using a Writer
transformer.  Or we check to make sure that all variables are in scope when
they're used -- using a Reader transformer.  We could even capture whitespace
and comment tokens, instead of just throwing them away, in case they contained
valuable information.  And the best part of it is, most of the parsers we already
wrote would continue to work fine without needing any modifications.  We'd just
have to set up our transformer stack and modify the relevant parsers!

>>>>

\subsection{Parser-specific combinators}
\verb+item+ is a very simple parser; it 
consumes a single token if available and fails otherwise:
\begin{verbatim}
item :: Parser t t
item = Parser f
  where 
    f [] = Nothing
    f (x:xs) = Just (xs, x)
\end{verbatim}

The \verb+check+ combinator is analogous to \verb+Control.Monad.mfilter+, which is used to
validate a parse result:
\begin{verbatim}
check :: (Monad f, Alternative f) => (a -> Bool) -> f a -> f a
check p m =
    m >>= \x -> if p x then return x else empty
\end{verbatim}

@@

Parsers built out of \verb+literal+ only succeed when the next token matches 
the specified one:
\begin{verbatim}
*Main> getParser (literal 'a') "abcde"
Just ("bcde",'a')

@@

*Main> getParser parseNest "(()abc(def)()"
Just ("(()abc(def)()",[])
\end{verbatim}
In the rest of this article, we'll rework and augment these basic parser 
combinators by generalizing the implementations and extending the result types,
while taking advantage of monad transformers to keep them modular.

\section{Monad transformers}
Monad transformers~\cite{liang} provide a method of combining monads into
new monads which support operations of each of their constituents.  
By applying monad transformers, we can build semantically rich parsers without sacrificing modularity.
We'll use the standard transformers/mtl \cite{mtl} libraries -- 
they're easy to install, easy to use, and very useful.

First, note that the parser type we used earlier -- \verb+[t] -> Maybe ([t], a)+
-- can be built out of the StateT monad transformer applied to Maybe.  \EZY{Do you maybe want to say why this is the case?}  We can 
now define the new parser type as:
\begin{verbatim}
import Control.Monad.State       (StateT(..))

@@

The \verb+item+ parser is reimplemented using the combinators from MonadState --
token lists are the `state' that \verb+item+ must inspect and modify:
\begin{verbatim}
{-# LANGUAGE   FlexibleContexts  #-}

@@

item :: (MonadState [t] m, Alternative m) => m t
item =
    get >>= \xs -> case xs of
                        (t:ts) -> put ts *> pure t;
                        []     -> empty;
\end{verbatim}
Note that the FlexibleContexts language extension is required by the 
MonadState constraint in the type signature.  This \EZY{which?} formulation of monad
transformers relies heavily on extensions to the type system, more of which 
we'll have to use for some of the later code in this article.

\section{Introducing Woof:  a Simple Lisp}

\EZY{An alternate structure you can use here is to introduce Woof first, before talking about monad transformers.  After all, the example parser doesn't require you to know what is going on in transformer-land.}

The simple language we'll use as a motivation for building parsers is Woof, a 
simple dialect of Lisp.  We'll be progressively adding features to a simple 
initial implementation on our way to creating an error-reporting parser.

@@

\subsection{Preliminaries}
Our first Woof parser will be responsible for determining whether input
text conforms to the language definition and building an 
Abstract Syntax Tree (AST) representing the structure of the parsed input.
Here's the AST definition we'll use:
\begin{verbatim}
data AST
    = ASymbol String
    | ALambda [String] [AST]
    | ADefine String AST
    | AApp    AST  [AST]
  deriving (Show, Eq)
\end{verbatim}

@@

And some examples to make sure it's behaving correctly: \EZY{An example?! Why not a QuickCheck property? \verb|;-)|}
\begin{verbatim}
*Main> map (runParser  opencurly) ["{abc", "}abc", "(abc", "abc"]
[Just ('{', "abc"), Nothing, Nothing, Nothing]
\end{verbatim}

@@

Of course not -- we haven't told it how to skip comments and whitespace yet!
Let's do that now, first by defining the whitespace and comment patterns, then
by creating a \verb+munch+ combinator that takes in a parser, 
throws away all leading junk, and finally runs the parser.  Also, we rename
the \verb+some+ and \verb+many+ functions from Control.Alternative to more
clearly indicate their semantics: \EZY{I would suspect many compiler authors would be of the opinion that the proper way to remove comments and whitespace is to lex first, and parse later.}
\begin{verbatim}
many0 = many
many1 = some

@@

On to Lambda!  Not only does Lambda have additional syntax with its parameter 
list, but we also need to ensure that the parameter names are unique.  We can 
do that using the \verb+check+ combinator again:
\begin{verbatim}
lambda = 
    opencurly                       *>
    check (== "lambda") symbol      *>
    opencurly                       *>
    pure ALambda                   <*>
    check distinct (many0 symbol)  <*>
    (closecurly                     *>
     many1 form                    <*
     closecurly)
  where
    distinct names = length names == length (nub names)
\end{verbatim}
If the symbols are distinct, the parameter list subparser will succeed,
whereas if there's a repeated symbol, it will fail.

@@

We also need to munch any trailing comments and whitespace -- otherwise \verb+end+
would fail -- so the final Woof parser is:
\begin{verbatim}
woof = many1 form <* munch end
\end{verbatim}

@@

While our previous parsers had two different results -- Nothing and Just --
our new parsers have three possible results:  success, failure, and error. \EZY{Question a reader might have: what is the difference between failure and error?  You give a little guidance later but give some intuition now!}  Note
that successful and failing parses are wrapped in an additional \verb+Right+
constructor:
\begin{verbatim}
*Main> runParser item "abc"
Right (Just ('a',"bc"))

@@

\subsection{Reporting position}
To report error position, we'll pre-process the input
string to calculate the line and column number of each character.  Then, instead 
of our parsers operating on a list of characters, they'll operate on a list of 
Tokens, where a Token is a character tupled up with a line number and column 
number, which we'll express with a simple typedef and some accessor functions:
\EZY{I don't think this is how most real parser libraries do it.  Why not do it
with some extra state on the parser monad; it's yet another application of the
transformer approach?}
\begin{verbatim}
type Token = (Char, Int, Int)
chr  (a, _, _)  =  a
line (_, b, _)  =  b
col  (_, _, c)  =  c
\end{verbatim}
The function responsible for calculating the position will simply transform a 
list of characters into a list of tokens:
\begin{verbatim}
countLineCol :: [Char] -> [Token]
countLineCol = reverse . snd . foldl f ((1, 1), [])
  where
    f ((line, col), ts) '\n' = ((line + 1, 1), ('\n', line, col):ts)
    f ((line, col), ts)  c   = ((line, col + 1), (c, line, col):ts)
\end{verbatim}

\subsection{Token parsers}
\EZY{In contrast, this section doesn't seem very enlightening, so maybe
it could be glossed over or relegated to an appendix.}
By adding the position to the input, we've created a new problem 
for ourselves: our original token parsers operated on 
character lists, but we now need them to work on token lists.  
The cause of the problem is the \verb+literal+ combinator, so we
replace it with a similar one for token lists:
\begin{verbatim}
character :: (MonadState [Token] m, 
              Alternative m) => Char -> m Token
character c = satisfy ((==) c . chr)
\end{verbatim}

Now our token parsers are:
\begin{verbatim}
whitespace = many1 $ satisfy (flip elem " \n\t\r\f" . chr)
comment = pure (:) <*> character ';' <*> many0 chars
  where chars = not1 $ character '\n'

@@

\subsection{Error messages}
Now, we modify the remaining parsers to report accurate and useful errors.
Here are the error messages that we'll use to report each error:
\begin{verbatim}
eAppOper    =  "application: missing operator"
eAppClose   =  "application: missing close parenthesis"
eDefSym     =  "define: missing symbol"
eDefForm    =  "define: missing form"
eDefClose   =  "define: missing close curly"
eLamParam   =  "lambda: missing parameter list"
eLamDupe    =  "lambda: duplicate parameter names"
eLamPClose  =  "lambda: missing parameter list close curly"
eLamBody    =  "lambda: missing body form"
eLamClose   =  "lambda: missing close curly"
eSpecial    =  "special form: unable to parse"
eWoof       =  "woof: unparsed input"
\end{verbatim}

The MonadError type class provides two useful combinators: \verb+throwError+ for 
generating errors and \verb+catchError+ for dealing with them.  We can ignore 
the second because we don't need to recover from errors, just report them, 
but when should an error be generated and what 
data should it include?  \verb+throwError+ is too basic for our needs, so we'll 
introduce the \verb+commit+ combinator:
\begin{verbatim}
commit :: (MonadError e m, Alternative m) => e -> m a -> m a
commit err p = p <|> throwError err
\end{verbatim}

This combinator takes two arguments: an error value and a monadic 
computation.  It first tries to run the computation; if the computation 
fails, it generates an error.  The 
definition of 'failure' is provided by the Alternative instance of \verb+m+.
\EZY{Give a little intuition on why the operator is called commit earlier;
you don't get around it until later.}

What data do we want to include in our errors?  A String describing the error 
and a token reporting the position will do:
\begin{verbatim}
type Error = (String, Token)
\end{verbatim}

@@

Note that \verb+commit+ isn't used until we can be sure that we're in the right 
rule: \EZY{Here it is}  if, in the \verb+define+ parser, we committed to parsing the ``define" 
keyword, we wouldn't be able to backtrack and try the \verb+lambda+ parser if 
that failed.  Thus, if errors are treated as unrecoverable, it's important not to 
place a \verb+commit+ where backtracking might be necessary to parse a valid 
alternative.  In our example, however, once 
we see an open curly brace and a define, we're sure that we're in the \verb+define+
rule; allowing useless backtracking would only destroy our ability to report 
errors cleanly and accurately.

@@

Incorrect input results in an error.  For each of the items in our error spec, 
we can show that our parser correctly reports an error: 
\begin{verbatim}
*Main> p "()"
Left ("application: missing operator",('(',1,1))

@@

\Red{It's worth noting that up until this point, we haven't talked about how transformers contribute to the design of the parser library.  Maybe give the juicy bits earlier.}

\section{Stack Order}
While the contents of a transformer stack is important, the order of the
layers is important as well~\cite{stack}.  Let's revisit our first Woof parser -- what if our 
stack had been Maybe/State instead of State/Maybe?
\begin{verbatim}
State/Maybe:   s -> Maybe (a, s)

@@

On the other hand, the order of Error and Maybe relative to each other isn't 
important.  But don't just take my word for it -- try it for yourself!
\EZY{It's worth mentioning why, formally speaking, this is the case, which has
to do with the relevant structures being isomorphic to each other.}

Another important point to note is that our parsers don't specify
which order -- Maybe/State or State/Maybe -- they prefer, and will 
work with either.  This means that the semantics are ultimately determined by 
the actual monad stack we use.

\section{Conclusion}
Monad transformers provide an elegant solution to error reporting in parser
combinators.  Better yet, they are extensible -- we could go back and add more
features to our parsers using additional transformers, perhaps extending our
parser to save partial results, so that if there's an error, it reports the
progress that it had made before the error.  We could do that using a Writer
transformer.  Or we check to make sure that all variables are in scope when
they're used -- using a Reader transformer.  We could even capture whitespace
and comment tokens, instead of just throwing them away, in case they contained
valuable information.  And the best part of it is, most of the parsers we already
wrote would continue to work fine without needing any modifications.  That
is the power of monad transformers!
